### 2023-02-10 chapters/abstract.tex
submitted version to the committee
with new formatting
fif edits fixed

\begin{abstract}
	\addcontentsline{toc}{chapter}{Abstract}
	\begin{center}
		Interpretability and Fairness in Machine Learning: A Formal Methods Approach\\
		\vspace{1em}
		by\\
		\vspace{1em}
		Bishwamittra Ghosh\\
		\vspace{1em}
	\end{center}

@@

	Fairness in machine learning centers on quantifying and mitigating the bias/unfairness of machine learning. In the presence of multiple fairness definitions and fairness algorithms, we propose a probabilistic fairness verifier, called {\justicia}, based on SSAT (stochastic satisfiability) with the goal of verifying whether the classifier achieves the desired level of fairness given the distribution of features. We also extend {\justicia} to consider feature correlations represented as a Bayesian Network, resulting in an accurate and scalable verification of fairness. Finally, we propose a framework for fairness influence functions (FIF) with an aim of quantifying the influence of features as their contribution on the bias of a classifier. FIF analysis interprets fairness by revealing potential features or subset of features attributing highly to the bias. Building on global sensitivity analysis, we propose an algorithm, called {\fairXplainer}, for computing FIFs of features, resulting in an accurate estimate of the bias of the classifier decomposed to input features.

>>>>

\begin{abstract}

@@

	Fairness in machine learning centers on quantifying and mitigating the bias or unfairness of machine learning. In the presence of multiple fairness definitions and fairness algorithms, we propose a probabilistic fairness verifier, called {\justicia}, based on SSAT (stochastic satisfiability) with the goal of verifying whether the classifier achieves the desired level of fairness given the distribution of features. We also extend {\justicia} to consider feature correlations represented as a Bayesian Network, resulting in an accurate and scalable verification of fairness. Finally, we propose a framework for fairness influence functions (FIF) with an aim of quantifying the influence of features as their contribution on the bias of a classifier. FIF analysis interprets fairness by revealing potential features or subset of features attributing highly to the bias. Building on global sensitivity analysis, we propose an algorithm, called {\fairXplainer}, for computing FIFs of features, resulting in an accurate estimate of the bias of the classifier decomposed to input features.

### 2023-02-10 chapters/background.tex
fif edits fixed

\subsection{Dataset and Distribution}
We consider a dataset $ \dataset $ as a collection of $n$ triples  $\{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n$ generated from an underlying distribution $\mathcal{D}$. Each non-sensitive data point $\mathbf{x}^{(i)}$ consists of $\numnonsensitive$ features $[x^{(i)}_1, \dots, x^{(i)}_{\numnonsensitive}] $. Each sensitive data point $\mathbf{a}^{(i)}$ consists of $\numsensitive$ categorical features $[a^{(i)}_1, \dots, a^{(i)}_{\numsensitive}] $.  $y^{(i)} \in \{0,1\}$ is the binary class corresponding to $(\mathbf{x}^{(i)}, \mathbf{a}^{(i)})$.

We use $ (\nonsensitive, \sensitive, Y) $ to denote the random variables corresponding to $ (\mathbf{x}, \mathbf{a}, y)$.  Each non-sensitive feature $ X_i \in \nonsensitive $ is sampled from a continuous probability distribution {$ \mathcal{X}_i $}, and each categorical sensitive feature $ A_j \in \sensitive $ is sampled from a discrete probability distribution {$ \mathcal{A}_j $}. Thus, $ \mathcal{D} $ is the product distribution of all features, $ \mathcal{D} \triangleq \prod_{i=1}^{\numnonsensitive}\mathcal{X}_i \prod_{j=1}^{\numsensitive} \mathcal{A}_j $.

@@

\paragraph{Equalized Odds ($ \mathsf{EO} $).} \textit{Separation} measuring group fairness metrics such as equalized odds constrain that $ \widehat{Y} $ is independent of $ \sensitive $ given the ground class $ Y $~\cite{hardt2016equality}.  Formally, for $ Y \in \{0,1\} $, equalized odds is

\begin{align*}
	f_{\mathsf{EO}}(\alg, \mathcal{D})  \triangleq \max(&\max_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}, Y = 0] - \min_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}, Y = 0], \\ &\max_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}, Y = 1] - \min_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}, Y = 1]).
\end{align*}

\paragraph{Predictive Parity ($ \mathsf{PP} $).} \textit{Sufficiency} measuring group fairness metrics such as predictive parity constrain that the ground class $ Y $ is independent of $ \sensitive $ given the prediction $ \widehat{Y} $~\cite{verma2018fairness}. Formally,

\begin{align*}
	f_{\mathsf{PP}}(\alg, \mathcal{D})  \triangleq \max(&\max_{\mathbf{a}}\Pr[Y =1 | \mathbf{A} = \mathbf{a}, \widehat{Y} = 0] - \min_{\mathbf{a}}\Pr[Y =1 | \mathbf{A} = \mathbf{a}, \widehat{Y} = 0], \\&\max_{\mathbf{a}}\Pr[Y =1 | \mathbf{A} = \mathbf{a}, \widehat{Y} = 1] - \min_{\mathbf{a}}\Pr[Y =1 | \mathbf{A} = \mathbf{a}, \widehat{Y} = 1]).
\end{align*}

@@

\blue{The complexity $ \bncomplexity(\graph) $ of a Bayesian network $ (\graph, \theta) $ with discrete random variables $ \graphvertices $ is defined as the number of independent parameters used to define the probability distribution of $ \graphvertices $~\cite{murakonda2021quantifying}. Let $ \bnnumdiscrete(\graphvertex) $ denote the number of distinct values that a random variable $ \graphvertex $ can take. As such, for each conditional probability $ \Pr[\graphvertex_i = v_i | \parent(\graphvertex_i) = \mathbf{u}; \theta] $, we need $ \bnnumdiscrete(\parent(\graphvertex_i))(\bnnumdiscrete(\graphvertex_i) - 1) $ independent parameters. Thus, the total complexity of a Bayesian network is:}

\blue{
\begin{align}
	\bncomplexity(\graph) = \sum_{i=1}^{|\graphvertices|} \bnnumdiscrete(\parent(\graphvertex_i))(\bnnumdiscrete(\graphvertex_i) - 1)
\end{align}
}

>>>>

\subsection{Dataset and Distribution}
We consider a dataset $ \mathbf{D} $ as a collection of $n$ data points  $\{(\mathbf{z}^{(i)}, y^{(i)})\}_{i=1}^n$  generated from an underlying distribution $\mathcal{D}$. The feature vector $\mathbf{z}^{(i)} \triangleq (\mathbf{x}^{(i)}, \mathbf{a}^{(i)}) $ is a concatenation of non-sensitive features $ \mathbf{x}^{(i)} $ and sensitive features $ \mathbf{a}^{(i)} $. Each non-sensitive data point $\mathbf{x}^{(i)}$ consists of $\numnonsensitive$ features $[x^{(i)}_1, \dots, x^{(i)}_{\numnonsensitive}] \in \mathbb{R}^{\numnonsensitive} $. Each sensitive data point $\mathbf{a}^{(i)}$ consists of $\numsensitive$ categorical features $[a^{(i)}_1, \dots, a^{(i)}_{\numsensitive}] \in \mathbb{N}^{\numsensitive} $. Thus, the cardinality of feature vector $ \mathbf{z}^{(i)} $ is denoted by $ m $, formally $ |\mathbf{z}^{(i)}| \triangleq \numfeatures = \numnonsensitive + \numsensitive $. The binary class corresponding to $(\mathbf{x}^{(i)}, \mathbf{a}^{(i)})$ is $y^{(i)} \in \{0,1\}$. We refer to $y^{(i)}$ as the true class.

We denote the random variables corresponding to $ (\mathbf{z}, \mathbf{x}, \mathbf{a},  y) $ with $ (\feature, \nonsensitive, \sensitive, Y) $. Each non-sensitive feature $ X_i \in \nonsensitive $ is sampled from a continuous probability distribution {$ \mathcal{X}_i $}, and each categorical sensitive feature $ A_j \in \sensitive $ is sampled from a discrete probability distribution {$ \mathcal{A}_j $}. Thus, $ \mathcal{D} $ is the product distribution of all features, $ \mathcal{D} \triangleq \prod_{i=1}^{\numnonsensitive}\mathcal{X}_i \prod_{j=1}^{\numsensitive} \mathcal{A}_j $.

@@

\paragraph{Equalized Odds ($ \mathsf{EO} $).} \textit{Separation} measuring group fairness metrics such as equalized odds~\cite{hardt2016equality} constrain that the predicted class $ \widehat{Y} $ is independent of $ \sensitive $ given the true class $ Y $.  Formally, for $ Y \in \{0,1\} $, equalized odds is defined as
\[ f_{\mathsf{EO}}(\alg, \mathbf{D}) \triangleq  \max_{y} \Big(\max_{\mathbf{a}}\Pr[\widehat{Y} =1 \mid \mathbf{A} = \mathbf{a}, Y = y] - \min_{\mathbf{a}}\Pr[\widehat{Y} =1 \mid \mathbf{A} = \mathbf{a}, Y = y]\Big).
\]

Equalized odds intuitively implies the maximum statistical parity conditioned on $ Y $.

\paragraph{Predictive Parity ($ \mathsf{PP} $).}
\textit{Sufficiency} measuring group fairness metrics such as Predictive Parity ($ \mathsf{PP} $) constrain that the ground class $ Y $ is independent of $ \sensitive $ given the prediction $ \widehat{Y} $. Formally, 
\begin{align*}
	f_{\mathsf{PP}}(\alg, \mathbf{D})  
	\triangleq \max_{y} \Big(\max_{\mathbf{a}}\Pr[Y =1 \mid \mathbf{A} = \mathbf{a}, \widehat{Y} = y] - \min_{\mathbf{a}}\Pr[Y =1 \mid \mathbf{A} = \mathbf{a}, \widehat{Y} = y]\Big).
\end{align*}

@@

The complexity $ \bncomplexity(\graph) $ of a Bayesian network $ (\graph, \theta) $ with discrete random variables $ \graphvertices $ is defined as the number of independent parameters used to define the probability distribution of $ \graphvertices $~\cite{murakonda2021quantifying}. Let $ \bnnumdiscrete(\graphvertex) $ denote the number of distinct values that a random variable $ \graphvertex $ can take. As such, for each conditional probability $ \Pr[\graphvertex_i = v_i | \parent(\graphvertex_i) = \mathbf{u}; \theta] $, we need $ \bnnumdiscrete(\parent(\graphvertex_i))(\bnnumdiscrete(\graphvertex_i) - 1) $ independent parameters. Thus, the total complexity of a Bayesian network is:

\begin{align}
	\bncomplexity(\graph) = \sum_{i=1}^{|\graphvertices|} \bnnumdiscrete(\parent(\graphvertex_i))(\bnnumdiscrete(\graphvertex_i) - 1)
\end{align}

### 2023-02-10 chapters/fairness/fif/appendix.tex
fif edits fixed

\appendix
\onecolumn
\section*{Appendix}
\section{Societal Impact}
In this paper, we quantify the influence of input features on the incurred bias/unfairness of the classifier on a dataset. This quantification facilitates our understanding of potential features or subsets of features attributing highly to the bias. This also allows us to understand the effect of the fairness enhancing algorithms in removing bias and their achievement/failure to do so. To the best of our understanding, this paper does not have any negative societal impact.

@@

\subsection{Equalized Odds}

\textit{Equalized Odds} ($ \mathsf{EO} $)~\cite{hardt2016equality}: \textit{Separation} measuring group fairness metrics such as equalized odds constrain that the predicted class $ \widehat{Y} $ is independent of $ \sensitive $ given the true class $ Y $.  Formally, for $ Y \in \{0,1\} $, equalized odds is defined as
\[ f_{\mathsf{EO}}(\alg, \mathbf{D}) \triangleq  \max_{y} \Big(\max_{\mathbf{a}}\Pr[\widehat{Y} =1 \mid \mathbf{A} = \mathbf{a}, Y = y] - \min_{\mathbf{a}}\Pr[\widehat{Y} =1 \mid \mathbf{A} = \mathbf{a}, Y = y]\Big).
\]

Equalized odds intuitively implies the maximum statistical parity conditioned on $ Y $.

\paragraph{Extension of {\fairXplainer} to Equalized Odds.}  For equalized odds, we deploy {\fairXplainer} twice, one for computing FIFs on a subset of data points in the dataset where $ Y = 1 $ and another on data points with $ Y = 0 $. Since, the maximum of the sum of FIFs between $ Y = 1 $ and $ Y = 0 $ is the equalized odds of the classifier, we finally report FIFs of features corresponding to the maximum sum of FIFs between $ Y \in \{0, 1\} $.

\subsection{Predictive Parity}

\textit{Sufficiency} measuring group fairness metrics such as Predictive Parity ($ \mathsf{PP} $) constrain that the ground class $ Y $ is independent of $ \sensitive $ given the prediction $ \widehat{Y} $. Formally, 
\begin{align*}
	f_{\mathsf{PP}}(\alg, \mathbf{D})  
	\triangleq \max_{y} \Big(\max_{\mathbf{a}}\Pr[Y =1 \mid \mathbf{A} = \mathbf{a}, \widehat{Y} = y] - \min_{\mathbf{a}}\Pr[Y =1 \mid \mathbf{A} = \mathbf{a}, \widehat{Y} = y]\Big).
\end{align*}

\paragraph{Extension of {\fairXplainer} to Predictive Parity.}  To compute FIFs for predictive parity, we condition the dataset by the predicted class $ \widehat{Y} $ and separate into two sub-datasets: $ \widehat{Y} = 1 $ and  $ \widehat{Y} = 0 $. For each sub-dataset, we deploy $ \fairXplainer $ by setting the ground-truth class $ Y $ as label. This contrasts the computation for  statistical parity and equalized odds, where the predicted class $ \widehat{Y} $ is considered as label. Finally, the maximum of the sum of FIFs between two sub-datasets for $ \widehat{Y} = 1 $ and  $ \widehat{Y} = 0 $ measures the predictive parity. Similar to equalized odds, FIFs achieving the greatest sum of FIFs for $ \widehat{Y} \in \{0, 1\} $ are the reported FIFs for the predictive parity of the classifier.

@@

\begin{table}
	\caption{Median estimation error (over 5-fold cross validation and all combinations of sensitive features) of equalized odds (columns $ 5 $ to $ 7 $) and predictive parity (columns $ 8 $ and $ 9 $) in terms of estimated FIFs by different methods. Best results (lowest error) are in bold color. `{\textemdash}' denotes timeout. SHAP cannot estimate FIFs for predictive parity due to limited methodology.}
	\label{fairness_fairXplainer_tab:accuracy_eo_pp}
	\centering
	%        \setlength{\tabcolsep}{.3em}
	\begin{tabular}{lrclrrr|rrr}
		\toprule
		\multirow{3}{*}{Dataset} & \multirow{3}{*}{\shortstack[1]{Dimension\\$ (n, \numfeatures) $}} & \multirow{3}{*}{\shortstack[1]{Max Sensitive\\Features, $ |\sensitive| $}} &\multirow{3}{*}{Classifier} & \multicolumn{3}{c|}{Equalized Odds} & \multicolumn{3}{c}{Predictive Parity} \\
		& & & & \multirow{2}{*}{SHAP} & \multicolumn{2}{c|}{\fairXplainer} &  \multicolumn{2}{c}{\fairXplainer}\\
		& & & & & $ \lambda = 1 $ & $ \lambda = 2 $ & $ \lambda = 1 $ & $ \lambda = 2 $ \\
		\midrule

@@

		\bottomrule
	\end{tabular}
\end{table}

@@

\begin{figure}
	\subfloat[Equalized Odds]{\includegraphics[scale=0.35]{figures/fairness/fif/runtime_diff_methods_train_eo}} 
	\subfloat[Predictive Parity]{\includegraphics[scale=0.35]{figures/fairness/fif/runtime_diff_methods_train_suff}}
	\caption{Execution time of different methods for estimating FIFs for equalized odds and predictive parity. {\fairXplainer} with $ \lambda = 1 $ is more efficient than SHAP, while {\fairXplainer} ($ \lambda = 2 $) requires more computational effort. SHAP cannot explain predictive parity. 
	}
	\label{fairness_fairXplainer_fig:execution_time_cactus_plot_eo_pp}
\end{figure}

@@

\begin{figure}
	\subfloat[Adult]{\includegraphics[scale=0.4]{figures/fairness/fif/sp_adult_train_error.pdf}}
	\subfloat[Adult]{\includegraphics[scale=0.4]{figures/fairness/fif/sp_adult_train_mean_rmse.pdf}}
	\subfloat[Adult]{\includegraphics[scale=0.4]{figures/fairness/fif/sp_adult_train_time.pdf}}

	\subfloat[COMPAS]{\includegraphics[scale=0.4]{figures/fairness/fif/sp_compas_train_error.pdf}}
	\subfloat[COMPAS]{\includegraphics[scale=0.4]{figures/fairness/fif/sp_compas_train_mean_rmse.pdf}}
	\subfloat[COMPAS]{\includegraphics[scale=0.4]{figures/fairness/fif/sp_compas_train_time.pdf}}

	\caption{Effect of spline intervals on the approximation error of statistical parity, root-mean square error (RMSE), and  execution time of {\fairXplainer}.}
	\label{fairness_fairXplainer_fig:effect_spline_intervals_appendix}
\end{figure}

@@

\begin{figure}
	\subfloat{\includegraphics[scale=0.4]{figures/fairness/fif/ablation_study_max_order_sp_train_error.pdf}}
	\subfloat{\includegraphics[scale=0.4]{figures/fairness/fif/ablation_study_max_order_sp_train_mean_rmse.pdf}}
	\subfloat{\includegraphics[scale=0.4]{figures/fairness/fif/ablation_study_max_order_sp_train_time.pdf}}

	\caption{Effect of maximum order $ \lambda $ on the approximation error of statistical parity, root-mean square error (RMSE) and  execution time of {\fairXplainer}.}
	\label{fairness_fairXplainer_fig:effect_maxorder_appendix}
\end{figure}

@@

\begin{figure}[!t]
	\begin{minipage}{0.45\textwidth}
		\centering
		\subfloat[Individual FIFs]{\includegraphics[scale=0.35]{figures/fairness/fif/adult_feature_weight_unsorted}\label{fairness_fairXplainer_fig:individual_fifs_adult}}
	\end{minipage}
	\begin{minipage}{0.5\textwidth}
		\centering
		\subfloat[Individual and intersectional FIFs]{\includegraphics[scale=0.35]{figures/fairness/fif/adult_feature_weight}\label{fairness_fairXplainer_fig:individual_and_intersectional_fifs_adult}}
	\end{minipage}
	\vspace{-0.2em}
	\caption{FIFs for Adult dataset on explaining statistical parity.}\label{fairness_fairXplainer_fig:individual_vs_intersectional_influence_adult}
\end{figure}

@@

\begin{figure}
	\centering
	\subfloat[Individual FIFs]{\includegraphics[scale=0.4]{figures/fairness/fif/titanic_feature_weight_unsorted}\label{fairness_fairXplainer_fig:individual_fifs_titanic}}\hfil
	\subfloat[Individual and intersectional FIFs]{\includegraphics[scale=0.4]{figures/fairness/fif/titanic_feature_weight}\label{fairness_fairXplainer_fig:individual_and_intersectional_fifs_titanic}}
	\caption{FIFs for Titanic dataset on explaining statistical parity.}\label{fairness_fairXplainer_fig:individual_vs_intersectional_influence_titanic}
\end{figure}

>>>>

\appendix
\chapter{Fairness Influence Functions}
\section{Societal Impact}
In this paper, we quantify the influence of input features on the incurred bias/unfairness of the classifier on a dataset. This quantification facilitates our understanding of potential features or subsets of features attributing highly to the bias. This also allows us to understand the effect of the fairness enhancing algorithms in removing bias and their achievement/failure to do so. To the best of our understanding, this paper does not have any negative societal impact.

@@

\subsection{FIFs of Equalized Odds.}  For equalized odds, we deploy {\fairXplainer} twice, one for computing FIFs on a subset of data points in the dataset where $ Y = 1 $ and another on data points with $ Y = 0 $. Since, the maximum of the sum of FIFs between $ Y = 1 $ and $ Y = 0 $ is the equalized odds of the classifier, we finally report FIFs of features corresponding to the maximum sum of FIFs between $ Y \in \{0, 1\} $.

\subsection{FIFs of Predictive Parity.}  To compute FIFs for predictive parity, we condition the dataset by the predicted class $ \widehat{Y} $ and separate into two sub-datasets: $ \widehat{Y} = 1 $ and  $ \widehat{Y} = 0 $. For each sub-dataset, we deploy $ \fairXplainer $ by setting the ground-truth class $ Y $ as label. This contrasts the computation for  statistical parity and equalized odds, where the predicted class $ \widehat{Y} $ is considered as label. Finally, the maximum of the sum of FIFs between two sub-datasets for $ \widehat{Y} = 1 $ and  $ \widehat{Y} = 0 $ measures the predictive parity. Similar to equalized odds, FIFs achieving the greatest sum of FIFs for $ \widehat{Y} \in \{0, 1\} $ are the reported FIFs for the predictive parity of the classifier.

@@

\begin{sidewaystable}
	\caption[Approximation error of $ \mathsf{EO} $ and $ \mathsf{PP} $ using FIFs]{Median estimation error (over 5-fold cross validation and all combinations of sensitive features) of equalized odds (columns $ 5 $ to $ 7 $) and predictive parity (columns $ 8 $ and $ 9 $) in terms of estimated FIFs by different methods. Best results (lowest error) are in bold color. `{\textemdash}' denotes timeout. SHAP cannot estimate FIFs for predictive parity due to limited methodology.}
	\label{fairness_fairXplainer_tab:accuracy_eo_pp}
	\centering
	%        \setlength{\tabcolsep}{.3em}
	\begin{tabular}{lrclrrr|rrr}
		\toprule
		\multirow{3}{*}{Dataset} & \multirow{3}{*}{\shortstack[1]{Dimension\\$ (n, \numfeatures) $}} & \multirow{3}{*}{\shortstack[1]{Max Sensitive\\Features, $ |\sensitive| $}} &\multirow{3}{*}{Classifier} & \multicolumn{3}{c|}{Equalized Odds} & \multicolumn{3}{c}{Predictive Parity} \\
		& & & & \multirow{2}{*}{SHAP} & \multicolumn{2}{c|}{\fairXplainer} &  \multicolumn{2}{c}{\fairXplainer}\\
		& & & & & $ \lambda = 1 $ & $ \lambda = 2 $ & $ \lambda = 1 $ & $ \lambda = 2 $ \\
		\midrule

@@

		\bottomrule
	\end{tabular}
\end{sidewaystable}

@@

\begin{figure}
	\centering
	\subfloat[Equalized Odds]{\includegraphics[scale=0.45]{figures/fairness/fif/runtime_diff_methods_train_eo}} 
	\subfloat[Predictive Parity]{\includegraphics[scale=0.45]{figures/fairness/fif/runtime_diff_methods_train_suff}}
	\caption[Execution time of FIFs]{Execution time of different methods for estimating FIFs for equalized odds and predictive parity. {\fairXplainer} with $ \lambda = 1 $ is more efficient than SHAP, while {\fairXplainer} ($ \lambda = 2 $) requires more computational effort. SHAP cannot explain predictive parity. 
	}
	\label{fairness_fairXplainer_fig:execution_time_cactus_plot_eo_pp}
\end{figure}

@@

\begin{figure}
	\centering
	\subfloat[Adult]{\includegraphics[scale=0.38]{figures/fairness/fif/sp_adult_train_error.pdf}}
	\subfloat[Adult]{\includegraphics[scale=0.38]{figures/fairness/fif/sp_adult_train_mean_rmse.pdf}}
	\subfloat[Adult]{\includegraphics[scale=0.38]{figures/fairness/fif/sp_adult_train_time.pdf}}

	\subfloat[COMPAS]{\includegraphics[scale=0.38]{figures/fairness/fif/sp_compas_train_error.pdf}}
	\subfloat[COMPAS]{\includegraphics[scale=0.38]{figures/fairness/fif/sp_compas_train_mean_rmse.pdf}}
	\subfloat[COMPAS]{\includegraphics[scale=0.38]{figures/fairness/fif/sp_compas_train_time.pdf}}

	\caption[Ablation study on FIFs: $ \tau $]{Effect of spline intervals on the approximation error of statistical parity, root-mean square error (RMSE), and  execution time of {\fairXplainer}.}
	\label{fairness_fairXplainer_fig:effect_spline_intervals_appendix}
\end{figure}

@@

\begin{figure}
	\centering
	\subfloat{\includegraphics[scale=0.4]{figures/fairness/fif/ablation_study_max_order_sp_train_error.pdf}}
	\subfloat{\includegraphics[scale=0.4]{figures/fairness/fif/ablation_study_max_order_sp_train_mean_rmse.pdf}}
	\subfloat{\includegraphics[scale=0.4]{figures/fairness/fif/ablation_study_max_order_sp_train_time.pdf}}

	\caption[Ablation study on FIFs: $ \lambda $]{Effect of maximum order $ \lambda $ on the approximation error of statistical parity, root-mean square error (RMSE) and  execution time of {\fairXplainer}.}
	\label{fairness_fairXplainer_fig:effect_maxorder_appendix}
\end{figure}

@@

\begin{figure}[!t]
	\begin{minipage}{0.45\textwidth}
		\centering
		\subfloat[Individual FIFs]{\includegraphics[scale=0.45]{figures/fairness/fif/adult_feature_weight_unsorted}\label{fairness_fairXplainer_fig:individual_fifs_adult}}
	\end{minipage}
	\begin{minipage}{0.5\textwidth}
		\centering
		\subfloat[Individual and intersectional FIFs]{\includegraphics[scale=0.45]{figures/fairness/fif/adult_feature_weight}\label{fairness_fairXplainer_fig:individual_and_intersectional_fifs_adult}}
	\end{minipage}
	\caption[FIFs for Adult dataset]{FIFs for Adult dataset on explaining statistical parity.}\label{fairness_fairXplainer_fig:individual_vs_intersectional_influence_adult}
\end{figure}

@@

\begin{figure}
	\centering
	\subfloat[Individual FIFs]{\includegraphics[scale=0.45]{figures/fairness/fif/titanic_feature_weight_unsorted}\label{fairness_fairXplainer_fig:individual_fifs_titanic}}\hfil
	\subfloat[Individual and intersectional FIFs]{\includegraphics[scale=0.45]{figures/fairness/fif/titanic_feature_weight}\label{fairness_fairXplainer_fig:individual_and_intersectional_fifs_titanic}}
	\caption[FIFs for Titanic dataset]{FIFs for Titanic dataset on explaining statistical parity.}\label{fairness_fairXplainer_fig:individual_vs_intersectional_influence_titanic}
\end{figure}

### 2023-02-10 chapters/fairness/fif/experiments.tex
fif edits fixed

\begin{figure}[t!]
	\centering
		\begin{minipage}{0.4\columnwidth}
		\centering
		\subfloat{\includegraphics[scale=0.38]{figures/fairness/fif/runtime_diff_methods_train_sp}}\vspace*{-.7em}
		\caption{Execution time of different methods for estimating FIFs. {\fairXplainer} with $ \lambda = 1 $ is more efficient than SHAP, while {\fairXplainer} ($ \lambda = 2 $) requires more computational effort.
		}
		\label{fairness_fairXplainer_fig:execution_time_cactus_plot}
	\end{minipage}\hfil
	\begin{minipage}{0.55\columnwidth}
		\centering\vspace*{-1em}
		\subfloat[COMPAS]{\includegraphics[scale=0.33]{figures/fairness/fif/intervention_compas}}
		\subfloat[Adult]{\includegraphics[scale=0.33]{figures/fairness/fif/intervention_adult}}\vspace*{-1em}
		\caption{Results on fairness intervention on logistic regression classifiers for COMPAS and Adult datasets. Pearson's correlation coefficient between bias-difference due to the intervention and FIFs is higher for top ranked influential features by {\fairXplainer} compared to SHAP.}	
		\label{fairness_fairXplainer_fig:fairness_intervention}
%		\centering
%		\subfloat{\includegraphics[scale=0.35]{figures/fairness/fif/sp_compas_train_error.pdf}}
%		\subfloat{\includegraphics[scale=0.35]{figures/fairness/fif/sp_compas_train_time.pdf}}
%		
%		\caption{Effect of spline intervals on the estimation error of statistical parity and  execution time of {\fairXplainer} on COMPAS.}
%		\label{fairness_fairXplainer_fig:effect_spline_intervals}
	\end{minipage}
\end{figure}

@@

	\caption{FIFs for COMPAS dataset on explaining statistical parity. Both individual and intersectional FIFs in Figure~\ref{fairness_fairXplainer_fig:intersectional_fifs} depict sources of bias in detail than individual FIFs alone in Figure~\ref{fairness_fairXplainer_fig:individual_fifs}. In Figure~\ref{fairness_fairXplainer_fig:fif_shap}, individual FIFs estimated by SHAP are far from correctly approximating statistical parity (exact value $0.174 $).}
	\label{fairness_fairXplainer_fig:fif_illustration_compas}
\end{figure}

@@

	\caption{Effects of a fairness attack~\cite{solans2020poisoning} and a fairness enhancing~\cite{kamiran2012data} algorithms on FIFs.}\label{fairness_fairXplainer_fig:affirmative_punitive_actions}
\end{figure}

>>>>

\begin{figure}
	\centering
	\subfloat{\includegraphics[scale=0.38]{figures/fairness/fif/runtime_diff_methods_train_sp}}
	\caption[Execution time of FIFs in $ \mathsf{SP} $]{Execution time of different methods for estimating FIFs. {\fairXplainer} with $ \lambda = 1 $ is more efficient than SHAP, while {\fairXplainer} ($ \lambda = 2 $) requires more computational effort.
	}
	\label{fairness_fairXplainer_fig:execution_time_cactus_plot}
\end{figure}
\begin{figure}
	\centering\vspace*{-1em}
	\subfloat[COMPAS]{\includegraphics[scale=0.33]{figures/fairness/fif/intervention_compas}}
	\subfloat[Adult]{\includegraphics[scale=0.33]{figures/fairness/fif/intervention_adult}}\vspace*{-1em}
	\caption[FIFs under fairness intervention]{Results on fairness intervention on logistic regression classifiers for COMPAS and Adult datasets. Pearson's correlation coefficient between bias-difference due to the intervention and FIFs is higher for top ranked influential features by {\fairXplainer} compared to SHAP.}	
	\label{fairness_fairXplainer_fig:fairness_intervention}
\end{figure}

@@

	\caption[FIFs for COMPAS dataset]{FIFs for COMPAS dataset on explaining statistical parity. Both individual and intersectional FIFs in Figure~\ref{fairness_fairXplainer_fig:intersectional_fifs} depict sources of bias in detail than individual FIFs alone in Figure~\ref{fairness_fairXplainer_fig:individual_fifs}. In Figure~\ref{fairness_fairXplainer_fig:fif_shap}, individual FIFs estimated by SHAP are far from correctly approximating statistical parity (exact value $0.174 $).}
	\label{fairness_fairXplainer_fig:fif_illustration_compas}
\end{figure}

@@

	\caption[FIFs under fairness affirmative/punitive actions]{Effects of a fairness attack~\cite{solans2020poisoning} and a fairness enhancing~\cite{kamiran2012data} algorithms on FIFs.}\label{fairness_fairXplainer_fig:affirmative_punitive_actions}
\end{figure}

### 2023-02-10 chapters/fairness/fif/figure_intro.tex
fif edits fixed

	\begin{minipage}[t]{0.07\textwidth}
		%\vspace{-2em}
		\scalebox{1}{	
			\begin{tikzpicture}[x=1cm,y=0.3cm]
			\node[] (a1) {\textbf{FIF:}};			
			\end{tikzpicture}
		}
	\end{minipage}%
	\begin{minipage}{0.46\textwidth}
		\centering
		\subfloat[Fairness influence functions (FIF) for DT$ 1 $]{\includegraphics[scale=0.42]{figures/fairness/fif/fif_example}
		\label{fairness_fairXplainer_fig:fif_original}}	
	\end{minipage}%
	\begin{minipage}{0.5\textwidth}
		\centering
		\subfloat[Modified FIFs for DT$ 2 $]{\includegraphics[scale=0.42]{figures/fairness/fif/fif_example_affirmative_action}
		\label{fairness_fairXplainer_fig:fif_affirmative}}
	\end{minipage}
%\vspace*{-.5em}
	\caption{FIFs of input features to investigate the bias (statistical parity) of a decision tree predicting the eligibility for health insurance using age-dependent features `fitness' and `income'. An affirmative action reduces bias as corresponding FIFs reflect it.}
	\label{fairness_fairXplainer_fig:fair_example_fif}
	%\vspace{-1.2em}
\end{figure}

>>>>

	\begin{minipage}[t]{0.07\textwidth}
		%\vspace{-2em}
		\scalebox{1}{	
			\begin{tikzpicture}[x=1cm,y=0.3cm]
			\node[] (a1) {\textbf{FIF:}};			
			\end{tikzpicture}
		}
	\end{minipage}%
	\begin{minipage}{0.46\textwidth}
		\centering
		\subfloat[Fairness influence functions (FIF) for DT$ 1 $]{\includegraphics[scale=0.42]{figures/fairness/fif/fif_example}
		\label{fairness_fairXplainer_fig:fif_original}}	
	\end{minipage}%
	\begin{minipage}{0.5\textwidth}
		\centering
		\subfloat[Modified FIFs for DT$ 2 $]{\includegraphics[scale=0.42]{figures/fairness/fif/fif_example_affirmative_action}
		\label{fairness_fairXplainer_fig:fif_affirmative}}
	\end{minipage}
%\vspace*{-.5em}
	\caption[Demonstration of FIF  in health insurance]{FIFs of input features to investigate the bias (statistical parity) of a decision tree predicting the eligibility for health insurance using age-dependent features `fitness' and `income'. An affirmative action reduces bias as corresponding FIFs reflect it.}
	\label{fairness_fairXplainer_fig:fair_example_fif}
	%\vspace{-1.2em}
\end{figure}
